Where Do Events Come From?

### 2018-08-02 paper.tex
...

\begin{abstract}
In reactive and event-based systems, execution is guided by an external
environment which generates inputs to the application and consumes outputs from
it.
Reactive languages provide dedicated syntax and semantics to deal with events
and greatly simplify the programming experience in this domain.
%
Nevertheless, the environment is typically prefabricated in a host language and
the very central concept of events is implemented externally to the reactive
language.
%
In this work, we propose an interrupt handler primitive for a reactive language
targeting embedded systems in order to take control of the whole event loop,
from a sensor input source and back to an actuator output.
%
We propose the new asynchronous primitive in the context of the synchronous
language \CEU and discuss how they synergize to avoid race conditions at
compile time, support lexically-scoped drivers, and provide automatic standby
for applications.
%and hot swapping of device drivers
\end{abstract}

@@

Reactive applications interact continuously and in real time with the external
world through sensors and actuators (e.g., buttons, displays, timers, etc.).
%
These interactions are typically represented as input events flowing from the
devices to the application and as output events flowing from the application to
the devices.
%
As illustrated in Figure~\ref{fig.env}, devices can be encapsulated as a
single component, \emph{the environment}, which controls the system execution
in an event loop:
the application sits idle waiting for an input;
the environment awakes the application on the occurrence of an input;
the application reacts to the input and generates back one or more outputs;
the application becomes idle and the loop restarts.

@@

The event-based interface between the application and the environment is
arguably inevitable and also happens at the right level of abstraction, since
it connects the application with the operating system resources through a
non-invasive API.
%
However, this separation exposes the environment as a rigid system component
that evolves in separate from the application.
%
It also requires two languages and the programmer has to deal with multiple
syntaxes, incompatible type systems, and different address spaces.
%
Furthermore, in the context of embedded systems, a proper host OS may even be
absent or lacking enough device drivers, which requires more low-level
intervention from the application.

In this work, we propose interrupt service routines (ISRs) as an asynchronous
primitive for the synchronous language \CEU~\cite{ceu.sensys13}.
ISRs empower applications to also implement their own device drivers and
self-generate inputs, bypassing the need for a host environment.
%
\CEU targets resource-constrained architectures, such as Arduino-compatible
microcontrollers, in which applications run in the bare metal, without
operating system support.
In this context, device drivers are typically libraries compiled together with
the main program.

The lack of an OS opens an opportunity to explore a tighter integration between
the application and its device drivers at the language level, resulting in an
overall simplicity and tractability of the system.
%
In particular, \CEU is amenable to a simple static analysis to detect race
conditions that we extended to encompass ISRs.
%
\CEU also provides a lexical finalization mechanism that we adopt in drivers to
properly disable interrupts.
%
Finally, the synchronous semantics of \CEU enforces that applications react
to inputs in bounded time and remain in an idle states susceptible to standby.
With the help of drivers and a power management runtime, \CEU can put the
microcontroller to sleep automatically at optimal sleeping modes after each
input reaction.

We implemented an extensible runtime that exposes hooks for the interrupts and
the power manager, which we validated in two microcontrollers: an
\emph{8-bit AVR/ATmega} and a \emph{32-bit ARM/Cortex-M0}.
%
We also implemented drivers for a variety of peripherals, such as GPIO, A/D
converter, USART, SPI, and nRF24L01 transceiver.
%
Applications can share data buffers with drivers to avoid unnecessary copying,
with the static guarantee that no data races will occur.
%
We demonstrate that applications built on top of these drivers show significant
energy savings due to automatic standby.

@@

what would be

output (int,int) O;

becomes

output (int x, int y) O do
end

@@

\CEU is a Esterel-based~\cite{ceu.tecs17} reactive programming language
targeting resource-constrained embedded systems~\cite{ceu.sensys13}, with the
characteristics that follow:
%
\begin{description}
\item [Reactive:] code only executes in reactions to events and is idle most of
    the time.
\item [Structured:] programs use lexically-scoped structured control
    mechanisms, such as \code{await} (to suspend a line of execution), and
    \code{par} (to combine multiple lines of execution).
\item [Synchronous:] reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{description}

@@

The example in Listing~\ref{lst.ceu} illustrates the main characteristics of
\CEU, namely event-driven I/O, lexically-scoped compositions and synchronous
execution.
%
The program toggles the state of the LED whenever a radio packet is received,
terminating on a button press, always with the LED off.
%
The program first declares the \code{LED} output, and the \code{BUTTON} and
\code{RADIO\_RECV} input events (ln~1--3).
The declarations include a payload type, i.e., each event occurrence carries
a value of that type (\code{high/low} is a boolean type).
%
Then, the program uses a \code{par/or} composition to run two lines of
execution, aka \emph{trails}, in parallel:
a single-statement trail that waits for a button press before terminating
(ln~5), and an endless loop that toggles the LED on and off whenever a radio
packet is received (ln~10--15).
Since we care about any reception occurrence, the example ignores the actual
packet payload (ln~12,14).
The \code{finalize} clause (ln~7--9) ensures that, no matter how its enclosing
trail terminates, the LED will be unconditionally turned off~(ln~8).

@@

The \code{par/or}, which stands for \emph{parallel-or}, is a lexical
composition that terminates as soon as one of the trails terminates, but which
also automatically finalizes the other trail(s).
%
In the example, when the button is pressed, not only the toggling loop will be
aborted, but the \code{finalize} clause will turn the LED off unconditionally,
since its enclosing block went out of scope.

The synchronous execution model of \CEU dictates that reactions to input events
are atomic and that incoming events are never lost.
%
The event loop in Figure~\ref{fig.env} indicates this behavior since the
environment can only generate a new input after the application yields control
and closes the loop.
%
In the example, even if two packets arrive simultaneously in the environment,
the synchronous model provides the atomicity and responsiveness guarantees:
    the first \code{await} will awake and turn the LED off atomically
    (ln~12--13);
    and the second \code{await} will awake in sequence (ln~14).
%
%Note that if multiple trails in parallel await the same input, they will all
%awake from it during the same reaction.
%In this case, \CEU dictates that each trail executes atomically, one after the
%other, in lexical order.
%This rule prevents race conditions for trails in parallel.

Figure~\ref{fig.ticks} illustrates the synchronous execution model of \CEU.
%
The continuous timeline in the environment shows an absolute reference clock
with ``physical timestamps'' for the event occurrences (e.g., event~\code{C}
occurs at 17ms521us).
%
The discrete timeline in the application shows how the same occurring events
fit in the logical notion of time of \CEU.
%
The boot reaction \code{boot-0} happens before any input, at program startup.
%
Event~\code{A} ``physically'' occurs during \code{boot-0} but, because time
is discrete, its corresponding reaction only executes afterwards, at logical
instant~\code{A-1}.
%
Similarly, event~\code{B} occurs during~\code{A-1} and its reaction is
postponed to execute at~\code{B-2}.
%
Event~\code{C} also occurs during~\code{A-1} but its reaction must also wait
for~\code{B-2} to execute and so it is postponed to execute at~\code{C-3}.
%
Finally, event~\code{D} occurs during an idle period and can start immediately
at \code{D-4}.
%
%Finally, two instances of event~\code{E} occur during~\code{D-4}; they are
%handled in the subsequent reactions~\code{E-5} and~\code{E-6}.

@@

>>>>

\begin{abstract}
In reactive and event-based systems, execution is guided by an external
environment which generates inputs to the application and consumes outputs from
it.
Reactive languages provide dedicated syntax and semantics to deal with events
and greatly simplify the programming experience in this domain.
%
Nevertheless, the environment is typically prefabricated in a host language and
the very central concept of events is implemented externally to the reactive
language.
%
In this work, we propose an interrupt handler primitive for a reactive language
targeting embedded systems in order to take control of the whole event loop:
from a sensor input source and back to an actuator output.
%
We propose the new asynchronous primitive in the context of the synchronous
language \CEU and discuss how they synergize to avoid race conditions at
compile time, support lexically-scoped drivers, and provide automatic standby
for applications.
%and hot swapping of device drivers
\end{abstract}

@@

Reactive applications interact continuously and in real time with the external
world through sensors and actuators peripherals (e.g., buttons, displays,
timers, etc.).
%
These interactions are typically represented as input events flowing from the
peripherals to the application and as output events flowing from the
application to the peripherals.
%
As illustrated in Figure~\ref{fig.env}, all peripherals can be encapsulated as
a single component, \emph{the environment}, which controls the system execution
in an event loop:
the application sits idle waiting for an input;
the environment awakes the application on the occurrence of an input;
the application reacts to the input and generates back one or more outputs;
the application becomes idle and the loop restarts.

@@

The event-based interface between the application and the environment is
arguably inevitable and also happens at the appropriate layer, since it
connects the application with the operating system resources through a
non-invasive API.
%
However, this separation exposes the environment as a rigid system component
that evolves in separate from the application.
%
It also requires two languages and the programmer has to deal with multiple
syntaxes, incompatible type systems, and different address spaces.
%
Furthermore, in the context of embedded systems, a proper host OS may even be
absent or lacking enough device drivers, which requires more low-level
intervention from the application.

In this work, we propose interrupt service routines (ISRs) as an asynchronous
primitive for the synchronous language \CEU~\cite{ceu.sensys13}.
ISRs empower reactive applications to also implement their own device drivers
and self-generate inputs, bypassing the need for a host environment.
%
\CEU targets resource-constrained architectures, such as Arduino-compatible
microcontrollers, in which applications run in the bare metal, without
operating system support.
In this context, device drivers are typically libraries compiled together with
the main program.

The lack of an OS opens an opportunity to explore a tighter integration between
the application and its device drivers at the language level, resulting in an
overall simplicity and tractability of the system.
%
In particular, \CEU is already amenable to a simple static analysis to detect
race conditions which we extended to also encompass ISRs.
Applications can share data buffers with drivers to avoid unnecessary copying,
with the static guarantee that no data races will occur.
%
\CEU also provides a lexical finalization mechanism that we adopt in drivers to
properly disable interrupts and turn off peripherals completely.
For instance, we can delimit the scope of a driver similarly to a
lexically-scoped variables.
%
Finally, the synchronous semantics of \CEU enforces that applications react
to inputs in bounded time and remain in an idle states susceptible to standby.
With the help of drivers and a power manager, we can put the microcontroller to
sleep automatically at optimal sleeping modes after each input reaction.

We implemented an extensible runtime that exposes hooks for the interrupts and
the power manager, which we validated in two microcontrollers: an
\emph{8-bit AVR/ATmega} and a \emph{32-bit ARM/Cortex-M0}.
%
We also implemented drivers for a variety of peripherals, such as GPIO, A/D
converter, USART, SPI, and nRF24L01 transceiver.
%
The applications built on top of these drivers show significant energy savings
due to automatic standby.

@@

@@

\CEU~\cite{ceu.sensys13} is a Esterel-based~\cite{ceu.tecs17} reactive
programming language targeting resource-constrained embedded systems, with the
characteristics that follow:
%
\begin{description}
\item [Reactive:] code only executes in reactions to events and is idle most of
    the time.
\item [Structured:] programs use lexically-scoped structured control
    mechanisms such as \code{spawn} and \code{await} (to create and suspend
    lines of execution).
\item [Synchronous:] reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{description}

@@

Listing~\ref{lst.ceu} illustrates the main characteristics of \CEU, namely
event-driven I/O, lexically-scoped compositions, and synchronous execution.
%
The program toggles the state of the LED whenever a radio packet is received,
terminating on a button press, always with the LED off.
%
The program first declares the \code{LED} output, and the \code{BUTTON} and
\code{RADIO\_RECV} input events (ln~1--3).
The declarations include a payload type, i.e., each event occurrence carries
a value of that type (\code{high/low} is a boolean type).
%
Then, the program uses a \code{par/or} composition to run two lines of
execution, aka \emph{trails}, in parallel:
a single-statement trail that waits for a button press before terminating
(ln~5), and an endless loop that toggles the LED on and off whenever a radio
packet is received (ln~10--15).
The \code{finalize} clause (ln~7--9) ensures that, no matter how its enclosing
trail terminates (e.g., from a button press), the LED will be unconditionally
turned off~(ln~8).

@@

The \code{par/or}, which stands for \emph{parallel-or}, is a lexical
composition that terminates as soon as one of the trails terminates, but which
also automatically finalizes the other trail(s).
%
In the example, when the button is pressed, not only the toggling loop will be
aborted, but the \code{finalize} clause will turn the LED off unconditionally,
since its enclosing block goes out of scope.

The synchronous execution model of \CEU dictates that reactions to input events
are atomic and that incoming events are never lost.
%
The event loop in Figure~\ref{fig.env} suggests this behavior since the
environment can only generate a new input after the application yields control
and closes the loop.
%
In the example, even if two packets arrive simultaneously in the environment,
the synchronous model ensures atomicity and responsiveness:
    the first \code{await} awakes and turns the LED off atomically
    (ln~12--13);
    and the second \code{await} will awake in sequence (ln~14).
%
%Note that if multiple trails in parallel await the same input, they will all
%awake from it during the same reaction.
%In this case, \CEU dictates that each trail executes atomically, one after the
%other, in lexical order.
%This rule prevents race conditions for trails in parallel.

Figure~\ref{fig.ticks} illustrates the synchronous execution model of \CEU.
%
The continuous timeline (representing the environment) shows
``physical timestamps'' for the event occurrences (e.g., event~\code{C}
occurs exactly at 17ms521us).
%
The discrete timeline in the application shows how the same occurring events
fit in the logical notion of time of \CEU.
%
The boot reaction \code{boot-0} happens before any input and starts the
program.
%
Event~\code{A} ``physically'' occurs during \code{boot-0} but, because time
is discrete, its corresponding reaction only executes at logical
instant~\code{A-1}.
%
Similarly, event~\code{B} occurs during~\code{A-1} and its reaction is
postponed to execute at~\code{B-2}.
%
Event~\code{C} also occurs during~\code{A-1} but its reaction must also wait
for~\code{B-2} to execute at~\code{C-3}.
%
Finally, event~\code{D} occurs during an idle period and can start immediately
at \code{D-4}.
%
%Finally, two instances of event~\code{E} occur during~\code{D-4}; they are
%handled in the subsequent reactions~\code{E-5} and~\code{E-6}.

@@

- TODO adc

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={XXX.},
  label={lst.gpio.async},
  escapechar=\%,
]
// A/D DRIVER (adc.ceu)

output int ADC_REQUEST;
input  int ADC_DONE;

do finalize with
  {
    ADCSRA &= B01111111;  // disables A/D converter
    ACSR    = B10000000;  // disables comparator
    DIDR0  |= B00111111;  // disables pins
  }
end

...     // driver implementation

loop do
    var int v;
    do
        #include "adc.ceu"
        emit ADC_REQUEST(A0);
        v = await ADC_DONE;
    end
    ...     // uses v
    await 1h;
end
\end{lstlisting}
